---
title: "Leo's Research Tools"
subtitle: "Guides, examples, and functions for statistical modeling"
format:
  html:
    toc: true
---

# Welcome

This website contains notes and examples for statistical models that I will inevitably forget how to run six months from now.

For our first example, let's investigate the relationship between speed and stopping distance using the built-in `cars` dataset.

# The Cars Data

```{r}
head(cars)
```

The dataset contains two variables:

-   `speed`: car speed in miles per hour
-   `dist`: stopping distance in feet

# Visualizing the Relationship

```{r}
plot(
  cars$speed,
  cars$dist,
  xlab = "Speed (mph)",
  ylab = "Stopping Distance (ft)",
  main = "Speed vs. Stopping Distance"
)
```

There appears to be a positive relationship between speed and stopping distance.

# Linear Regression

We can estimate this relationship using a simple linear regression:

$$
Distance_i = \beta_0 + \beta_1 Speed_i + \epsilon_i
$$

```{r}
mod <- lm(dist ~ speed, data = cars)

summary(mod)
```

# Regression Plot

```{r}
plot(
  cars$speed,
  cars$dist,
  xlab = "Speed (mph)",
  ylab = "Stopping Distance (ft)"
)

abline(mod)
```

# Conclusion

Cars traveling at higher speeds tend to require greater stopping distances.

Eventually, this page will instead contain links to guides covering topics such as EMC2, diffusion decision models, IRT, mixed-effects models, and simulation-based power analysis.
