---
title: "Introduction to Go/No-Go Wiener Diffusion Models"
editor: visual
---

In this document, we will be learning how to simulate Go/No-Go (GNG) data directly rather than transforming simulated two-choice data into a GNG. Additionally, we will be discussing how to specify GNG specific prior distributions from the existing literature, how to fit a GNG WDM, check whether the model converged, assess if the model does a good job explaining the data and then running a parameter recovery test. 


```{r load in packages, warning = F, message= F}
# load in packages
library(here)
library(tidyverse)
library(ggplot2)
library(EMC2)
```

```{r load in custom functions}
# load in all custom ddm functions
source(here("ddm", "load.R"))
```

## 1. Correctly simulating GNG data

Like in the previous guides, we will only be fitting **Wiener diffiusion models** (WDM), which is a DDM where between-trial variability parameters (e.g., Sz, St0, SZ) are not estimated. Thus our model objects (emc objects to be specific) will have 'WDM' in the name.

### 1.1 Specifying a GNG WDM

We start by creating a model design object ('emc.design') using the `design()` function where we specify that we will be fitting a 'DDMGNG', what the factors of our dataset will be (e.g., esubjects, S), what the responses will be ("TO" is the reference so it will map to the lower boundary), our WDM specifications within formula, and additional information about what values within R correspond to a go vs a no-go response.

```{r Creating a GNG design object}
#| output: false

# Generating the model specifications for a GnG dataset
designWDMGNG <- design(model = DDMGNG,
                       factors=list(subjects=1, S=c("ng", "go")),
                       Rlevels = c("TO", "RESP"),
                       formula = list(a ~ 1,
                                      v ~ S, 
                                      t0 ~ 1,
                                      Z ~ 1),
                       functions = list(
                         TIMEOUT = function(d) rep(1.5, nrow(d)),
                         Rnogo  = function(d) factor(rep("TO",   nrow(d)), levels = c("TO", "RESP")),
                         Rgo    = function(d) factor(rep("RESP", nrow(d)), levels = c("TO", "RESP"))
                        ))
```

### 1.2 Choosing the GNG WDM parameter values

Here we use the `sampled_pars()` function to extract an empty vector of parameter names that will be estimated. We then choose what these values will be and confirm that we did this correctly with the `mapped_pars()` function.

```{r Choosing the WDM parameter values}
# Choose parameters for simulation, first getting a vector with the right name
parsWDMGNG <- sampled_pars(designWDMGNG)

# Then choosing some values. Note that we include transforms here to illustrate
# the scale on which sampling occurs
parsWDMGNG[] <- c(log(1), -1, 4.2, log(.3), qnorm(.5))

# Confirm our parameters were specified correctly
mapped_pars(designWDMGNG, parsWDMGNG)
```

### 1.3 Simulating Racliff et al. (2018) GNG data

We are using Ratcliff et al. (2018) same data structure from their study to generate our GnG data. Their GNG task had 960 trials for each subject, where 720 trials were 'go' trials (75%) and 240 trials (25%) were 'no-go' trials. Thus we will do the same but just for one subject, as we will not be fitting hierarchical models in this document.

When simulating the data, we must make sure that all of our non-numeric variables are factors and not characters, and we must make the reference level consistent with the emc model object, such that "ng" and "TO" are the reference levels.

```{r making the data skeleton structure}
# Specify task structure
n_trials <- 960
prop_nogo <- 0.25

# Calculate number of Go and No-Go trials
n_nogo <- round(n_trials * prop_nogo)
n_go   <- n_trials - n_nogo

# Create simulation framework
GNG_framework <- data.frame(
  trials = seq_len(n_trials),
  subjects = factor(1),
  S = factor(
    sample(c(rep("go", n_go), rep("ng", n_nogo))),
    levels = c("ng", "go")
  ),
  R = factor(
    rep(NA_character_, n_trials),
    levels = c("TO", "RESP")
  )
)

# Simulate the gng data
gng <- make_data(data = GNG_framework,
                 parameters = parsWDMGNG, 
                 design = designWDMGNG) 
```

### 1.4 Getting descriptives from the GNG data

We can use the `data_design()` custom function to be able to use the `get_descriptives()` function to get some nice descriptives of our GNG data. We can see the number of trials, the proportion of which are correct within stimulus type ('ng', 'go') and information on reaction times.

```{r creating the data_design object and getting descriptives}
# Create the data_design for the gng data
gng_design <- data_design(
  stimulus = S,
  response = R,
  rt = rt,
  correct = (S == "go" & R == "RESP") | (S == "ng" & R == "TO"),
  id = subjects
)

# Get the descriptives from the datasets
get_descriptives(gng, gng_design)

```

### 1.5 Plotting the defective density distribution for a GNG

Here we can use the `plot_defectiveDensityGNG()` function to compare the reaction time distribution between 'go' (correct) and 'ng' (incorrect) trials.

```{r Plotting a defective density for the gng data}
# Plot the defective density info
plot_defectiveDensityGNG(gng, gng_design)
```

## 2. Generating GNG Specific Prior Distributions

### 2.1 Specifying prior means for a GNG

To come up with theoretical values for the GNG DDM, we identified a four papers that have used a GNG DDM, along with their reported results. **Note**: in the columns, Z is the relative starting point which is defined as z/a. Additionally, the following table was generated using a mixture of help from Notebook LM and ChatGPT.

| Study | Sample | Model | Scaling | Go v | NG v | a | Ter | Z |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| Gomez et al. (2007) Exp 1 | Healthy young adults (n = 20) | GnG DDM | s = 0.1 | Med-Freq: 0.279, Low-Freq: 0.147 | Nonword: -0.240 | a = 0.125 | Ter = 0.436s | Z = 0.544 |
| Gomez et al. (2007) Exp 2 | Healthy young adults (n = 19) | GnG DDM | s = 0.1 | Med-Freq: 0.259, Low-Freq: 0.114 | Nonword: -0.245 | a = 0.101 | Ter = 0.411s | Z = 0.604 |
| Gomez et al. (2007) Exp 3 | Healthy young adults (n = 21) | GnG DDM | s = 0.1 | Nonword Go: 0.207 | Med-Freq: -0.215, Low-Freq: -0.096 | a = 0.114 | Ter = 0.474s | Z = 0.474 |
| Gomez et al. (2007) Exp 4 | Healthy young adults (n = 28) | GnG DDM | s = 0.1 | Accuracy: 0.262 (Med). 0.096 (Low), Speed: 0.262 | Accuracy: -0.265, Speed: -0.265 | a = 0.105 - 0.114 | Ter = 0.435s (both) | Z = 0.570 - 0.657 |
| Gomez et al. (2007) Exp 5 | Healthy young adults (n = 11) | GnG DDM | s = 0.1 | Easy: 0.591, Hard: 0.178 | Easy: -0.404, Hard: -0.009 | a = 0.111 | Ter = 0.263s | Z = 0.577 |
| Gomez et al. (2007) Exp 6 | Healthy young adults (n = 14) | GnG DDM | s = 0.1 | v1: 0.136, v2: 0.252 | v1: -0.220, v2: -0.262 | a = 0.123 | Ter = 0.367s | Z = 0.496 |
| Ratcliff et al. (2018) Exp 1 | Healthy young adults (n=39) | DDM GnG (75% Go) | s = 0.1 | easy: 0.341, hard: 0.223 | easy: -0.162, hard: -0.008 | a = 0.139 | Ter = 0.379 | Z = 0.608 |
| Ratcliff et al. (2018) Exp 2 | Healthy young adults (n=31) | DDM GnG (75% Go) | s = 0.1 | easy: 0.294, hard: 0.202 | easy: -0.133, hard: -0.011 | a = 0.145 | Ter = 0.363 | Z = 0.523 |
| Huang-Pollock et al. (2017) | Children (8-12), ADHD: n = 46, Control: n = 21 | GnG DDM (75% Go) | s = 0.1 | ADHD = 0.238, Ctrl = 0.285 | ADHD = -0.157, Ctrl = -0.197 | ADHD = 0.162, Ctrl = 0.160 | ADHD = 0.415s, Ctrl = 0.437s | ADHD = 0.568, Ctrl = 0.625 |
| Huang-Pollock et al. (2017) | Children (8-12), ADHD: n = 51, Control: n = 18 | GnG DDM (75% Go) | s = 0.1 | ADHD = 0.223, Ctrl = 0.250 | ADHD = -0.109, Ctrl = -0.198 | ADHD = 0.152, Ctrl = 0.152 | ADHD = 0.393s, Ctrl = 0.403s | ADHD = 0.678, Ctrl = 0.691 |
| Cao et al. (2026) | Healthy adults (N = 36) | DDM GNG | s = 1.0 | go: 4.093 | ng: -2.194 | a: 1.579 | Ter = 0.253s | Z = 0.402 |
| Cao et al. (2026) | Healthy adults (N = 36) Sleep Deprived | DDM GNG | s = 1.0 | go: 2.947 | ng: -1.670 | a: 0.990 | Ter = 0.256s | Z = 0.462 |

We can use the information above to generate a **broad summary table** to better make sense of this information. These parameter estimates are **means**, meaning they are an average of multiple participant's GNG performance. Due to this, while not present below, it is possible for **no-go drift rates to be positive** due to poor performance.

| Parameter | Estimate (n) | Mean   | SD     | Min    | Max    |
|-----------|--------------|--------|--------|--------|--------|
| Go v      | 22           | 2.520  | 1.053  | 0.960  | 5.910  |
| Ng v      | 20           | -1.791 | 0.992  | -4.040 | -0.080 |
| a         | 15           | 1.307  | 0.229  | 0.990  | 1.620  |
| Ter       | 15           | 0.381s | 0.071s | 0.253s | 0.474s |
| Z         | 15           | 0.565  | 0.083  | 0.402  | 0.691  |

From this information, we can essentially now create prior distributions where we use the means presented here along with the standard deviations but in the transformed scale.

**Citations**:

-   Ratcliff, R., Huang-Pollock, C., & McKoon, G. (2018). Modeling individual differences in the go/no-go task with a diffusion model. Decision, 5(1), 42.
-   Huang-Pollock, C., Ratcliff, R., McKoon, G., Shapiro, Z., Weigard, A., & Galloway-Long, H. (2017). Using the diffusion model to explain cognitive deficits in attention deficit hyperactivity disorder. Journal of abnormal child psychology, 45(1), 57-68.
-   Gomez, P., Ratcliff, R., & Perea, M. (2007). A model of the go/no-go task. Journal of Experimental Psychology: General, 136(3), 389.
-   Cao, Y., Luo, J., Li, J., Wang, L., & Ma, N. (2026). Commonality and diversity of impairments in inhibitory tasks after total sleep deprivation: evidence from drift diffusion model. SLEEPJ, 49(3), zsaf355.

```{r specifying the GNG prior means}
# Specify the prior means in the transformed scale
pmean <- c(
  a = log(1.31), 
  v = -1.8, 
  v_Sgo = 2.5,  
  t0 = log(.38),
  Z = qnorm(.57) 
)
```

### 2.2 Specifying prior standard deviations for a GNG

In the introduction to the DDM document, we used the following rule to calculate prior standard deviations (a heuristic):

$$ SD = \frac{\text{MAX} - \text{MIN}}{4}$$ The problem is, when we did that we had estimates nicely summarized for us from a study that reviewed 23 DDM studies. In our case, we reported parameter estimates from literally 4 studies- thus using the formula above, while informative, might lead us to have very restricted ranges. Instead, what we can do is keep the same logic but change the denominator from 4 down to 2 for now, until there are more studies available or someone nicely summerizes it all for us.

$$ SD_{GNG} = \frac{\text{MAX} - \text{MIN}}{2}$$

boundary separation (a): (log(1.620) - log(0.990))/2 = 0.25 non-decision time (Ter): (log(0.474) - log(0.253))/2 = 0.31 relative-starting point (Z): (qnorm(0.691) - qnorm(0.402))/2 = 0.37

We purposly kept the drift rates wide (2.5) rather than using the formula above, just in case we have very poor performers in our datasets. The same decision was made when specifying the prior distribution for the 2C DDM.

```{r Specifying the GNG prior standard deviations}
# Standard deviations for the GNG priors
psd <- c(
  a = 0.25,     # Spread for log(a)
  v = 2.5,     # Spread for drift rate v (ng)
  v_Sgo = 2.5,  # Spread for drift rate v (go)
  t0 = 0.31,   # Spread for log(t0)
  Z = 0.37     # Spread for probit(Z)
)

# Standard deviations from the 2C DDM priors
#psd <- c(
#  a = 0.5,     # Spread for log(a)
#  v = 2.5,     # Spread for drift rate v (S1)
#  v_S2 = 2.5,  # Spread for drift rate v (S2)
#  t0 = 0.38,   # Spread for log(t0)
#  Z = 0.35     # Spread for probit(Z)
#)
```

### 2.3 Creating the GNG prior distributions

With the means and standard deviations coded for, we can now use the `prior()` function to generate our priors. Additionally, we also specify that this is for a non-hierarchical model.

```{r Creating the GNG prior distributions}
# Creating the prior distribution for each GNG WDM parameter
priorWDMGNG <- prior(designWDMGNG,
                     pmean=pmean, 
                     psd=psd,
                     type="single") # Non-hierarchical

class(priorWDMGNG)

# Print out the prior information (inspect for anything that remains at the default)
priorWDMGNG
```

### 2.4 Plotting the GNG prior distributions

Here we can visualize the prior distributions just to confirm that everything looks okay.

```{r plotting the GNG prior distributions}
# plotting on the natural scale
plot(priorWDMGNG, designWDMGNG, layout=c(2,4))
```

## 3. Fitting a GNG WDM Model

### 3.1 Making an emc object and fitting the GNG WDM

Before we can fit the model, we need to create an emc object using the `make_emc()` function. For this, we input the data, the model design, the prior distribution and mention that we are fitting a non-hierarchical model. Afterwards, we can use the `fit()` function to run the model.

```{r Fitting the model}
# Create an emc object for the gng data
sWDM_GNG2 <-  make_emc(data = gng, 
                       design = designWDMGNG, 
                       type= "single",
                       prior=priorWDMGNG)

# Fitting the model
# sWDM_GNG2 <- fit(sWDM_GNG2)

# Saving the model objects
# save(sWDM_GNG2, file = here("ddm", "individual", "models", "sWDM_GNG2.Rdata"))

# Loading the model objects
load(here("ddm", "individual", "models", "sWDM_GNG2.Rdata"))
```

### 3.2 Checking model convergence

From the output below, the model seems to converge just fine. All Rhat values are less than 1.1 and the ESS values are all greater than 1,000.

```{r Checking model convergence}
# Check model convergence
check(sWDM_GNG2)
```

### 3.3 Extracting model parameters

Below we can use the `mapped_pars()` function to view the parameters that were initially chosen when simulating the data. The `credict()` function shows us that the GNG WDM was able to nicely estimate these same parameters!

```{r extracting the model parameters}
# The original specified parameters (ground truth)
mapped_pars(designWDMGNG, parsWDMGNG)

# The estimated parameters from the GNG WDM
credint(sWDM_GNG2, selection = "alpha", map = TRUE)
```

## 4. Does the GnG WDM capture the data well?

### 4.1 Producing the predicted values

To assess whether our model can explain our data well, we can use the `predict()` function to produce 50 predicted datasets. Each one is the same dimension as our original GNG dataset and contains the same number of trials for each stimuli. Essentially. we can use these datasets to generate an 'average' predicted dataset, made up by values in the 50th percentile and also create a 95% credible interval using values at the 2.5 and 97.5 percentiles.

```{r Getting the predicted values}
# Use the predict function to obtain the fitted values
# ppWDM_GNG2 <- predict(sWDM_GNG2)

# Saving the predicted object
# saveRDS(ppWDM_GNG2, file = here("ddm", "individual", "predicted_values", "ppWDM_GNG2.rds"))

# Loading the predicted values data frame
ppWDM_GNG2 <- readRDS(here("ddm", "individual", "predicted_values", "ppWDM_GNG2.rds"))

# Why so many rows?
xtabs(~S+ postn, ppWDM_GNG2)
```

### 4.2 Plotting the predicted values over the GNG dataset

Below we use the `plot_density()` and `plot_cdf()` functions to produce defective distributions for the go and ng stimuli. Both plots are essentially saying the same thing. Overall, visually it appears that the original data fits between the 95% credible interval created by the predicted datasets. Moreover, we see these confidence intervals are much large for the ng trials than the go trials. This is just a consequence of how many trials are available in the dataset.

```{r plotting the predicted values}
# plotting predicted values as a density plot
plot_density(gng, ppWDM_GNG2, factors = c("S"))

# Plotting the predicted values as a cumulative defective function
plot_cdf(gng, ppWDM_GNG2, factors = c("S"))
```

### 4.3 Quickly checking accuracy

We can use the `aggregate()` function below to quickly gauge how well the predicted datasets are capturing accuracy within their 95% credible interval- overall it looks good!

```{r Getting descriptives of the original data}
# Getting descriptives for the original data
agg1 <- aggregate(R=="RESP"~subjects*S, gng, mean)
colnames(agg1)[3] <- c("RESP")
agg2 <- aggregate(RESP~S, agg1, mean)

# Getting descriptives for the predicted datasets
agg1_pp <- aggregate(R=="RESP"~ S*subjects*postn, ppWDM_GNG2, mean)
colnames(agg1_pp)[4] <- c("RESP")
agg2_pp <- aggregate(RESP~S*postn, agg1_pp, mean)
agg3_pp <- aggregate(RESP~S, agg2_pp, median)
low_pp <- aggregate(RESP~S, agg2_pp, function(x) quantile(x, 0.025))
upp_pp <- aggregate(RESP~S, agg2_pp, function(x) quantile(x, 0.975))

agg4_pp <- data.frame(
  S      = agg3_pp$S,
  "2.5%"  = low_pp$RESP,
  "50%"   = agg3_pp$RESP,
  "97.5%" = upp_pp$RESP,
  check.names = FALSE  
)

# Raw data vs Predicted datasets
agg2; agg4_pp
```

## 5. Parameter Recovery

For brevity below we do a parameter recovery test and it looks splendid.

```{r Parameter Recovery}
# Simulate data from our model parameters
sim <- make_data(parameters = sWDM_GNG2,
                 design = designWDMGNG,
                 data=gng)

# Make an emc object
sim_sWDM_GNG2 <- make_emc(data = sim,
                          design = designWDMGNG,
                          type="single",
                          prior = priorWDMGNG)

# Fit the model to the simulated data
# sim_sWDM_GNG2 <- fit(sim_sWDM_GNG2)

# Saving the model object
# save(sim_sWDM_GNG2, file = here("ddm", "individual", "models", "sim_sWDM_GNG2.Rdata"))

# Loading the model object
load(here("ddm", "individual", "models", "sim_sWDM_GNG2.Rdata"))

# Save the original parameters from the model (in transformed scale)
ground_truth_params <- credint(sWDM_GNG2)[[1]][,"50%"]

# Compare parameters from the original to the simulated data
recovery(emc = sim_sWDM_GNG2, 
         true_pars = ground_truth_params)
```
