---
title: "Fitting Between-Trial Variability Parameters to GNG data"
editor: visual
---

```{r load in packages, warning = F, message= F}
# load in packages
library(here)
library(tidyverse)
library(ggplot2)
library(EMC2)
```

```{r load in custom functions}
# load in all custom ddm functions
source(here("ddm", "load.R"))
```

## 1. Read in the data

We will be reading the already processed and combine GNG datasets.

```{r reading in the data}
# Read in the data
gng_combined<- readRDS(here("ddm", "individual", "data", "gng_combined.rds"))

# View the structure of the data
str(gng_combined)
```

## 2. Specifying Between-Trial Variability DDMs

### 2.1 Creating a custom function to generate between-trial variability DDMs

```{r Making a custom function to help with creting between trial variability DDMs}
# Create a custom function that will help in creating multiple between-trial variability DDM
make_designDDMGNG <- function(formula, LT = .250) {
  
  design(
    model = DDMGNG, 
    factors = list( subjects = 1, S = c("ng", "go"), E = c("spd", "acc") ), 
    Rlevels = c("TO", "RESP"), 
    LT = LT, 
    formula = formula, 
    functions = list(
      TIMEOUT = function(d) rep(1.5, nrow(d)), 
      Rnogo = function(d) factor(rep("TO", nrow(d)), levels = c("TO", "RESP")), 
      Rgo = function(d) factor(rep("RESP", nrow(d)), levels = c("TO", "RESP")))
  )
}
```

### 2.2 Creating the between-trial variability DDM designs

```{r Creating the between trial variability DDM designs}
#| output: false

# Create a formula object with the WDM parameter specification
WDM_formula <- list(a ~ E, 
                    v ~ S * E, 
                    t0 ~ E, 
                    Z ~ E)

# Specifying GNG DDM designs for estimating between-trial variability parameters (single)
designGNG_SZ <- make_designDDMGNG(formula = c(WDM_formula, list = (SZ ~ 1)))
designGNG_sv <- make_designDDMGNG(formula = c(WDM_formula, list = (sv ~ 1)))
designGNG_st0 <- make_designDDMGNG(formula = c(WDM_formula, list = (st0 ~ 1)))

# Specifying GNG DDM designs for estimating between-trial variability parameters (multiple)
designGNG_SZ_sv <- make_designDDMGNG(formula = c(WDM_formula, list(SZ ~ 1, sv ~ 1)))
designGNG_SZ_st0 <- make_designDDMGNG(formula = c(WDM_formula, list(SZ ~ 1, st0 ~ 1)))
designGNG_sv_st0 <- make_designDDMGNG(formula = c(WDM_formula, list(sv ~ 1, st0 ~ 1)))

# Specifying the full GNG DDM
designGNG_full <- make_designDDMGNG(formula = c(WDM_formula, list(SZ ~ 1, sv ~ 1, st0 ~ 1)))
```

### 2.3 Creating priors distributions

```{r creating prior distributions}
# Prior means on EMC2's transformed scale
pmean <- c(
  a       = log(1.31),
  v       = -1.8,
  v_Sgo   = 2.5,
  t0      = log(.38),
  Z       = qnorm(.57),

  # Between-trial variability parameters
  SZ      = qnorm(.38),
  sv      = log(1.2),
  st0     = log(.03)
)

# Prior SDs on EMC2's transformed scale
psd <- c(
  a       = 0.25,
  v       = 2.5,
  v_Sgo   = 2.5,
  t0      = 0.31,
  Z       = 0.37,

  # Between-trial variability parameters
  SZ      = 0.75,
  sv      = 0.70,
  st0     = 0.90
)

# Creating the specific prior distributions for each between-trial variability DDM
priorGNG_SZ <- prior(designGNG_SZ, pmean = pmean, psd = psd, type = "single")
priorGNG_sv <- prior(designGNG_sv, pmean = pmean, psd = psd, type = "single")
priorGNG_st0 <- prior(designGNG_st0, pmean = pmean, psd = psd, type = "single")

priorGNG_SZ_sv <- prior(designGNG_SZ_sv, pmean = pmean, psd = psd, type = "single")
priorGNG_SZ_st0 <- prior(designGNG_SZ_st0, pmean = pmean, psd = psd, type = "single")
priorGNG_sv_st0 <- prior(designGNG_sv_st0, pmean = pmean, psd = psd, type = "single")

priorGNG_full <- prior(designGNG_full, pmean = pmean, psd = psd, type = "single")
```

```{r Printing the default priors for the full GNG}
# Print the full prior information for the full GNG
priorGNG_full
```

### 2.4 Creating the emc objects

```{r Creating the emc objects}
# Create the emc objects so we can fit the DDMs (single)
sDDM_GNG_SZ <- make_emc(data = gng_combined, 
                        design = designGNG_SZ, type="single", prior=priorGNG_SZ)
sDDM_GNG_sv <- make_emc(data = gng_combined, 
                        design = designGNG_sv, type="single", prior=priorGNG_sv)
sDDM_GNG_t0 <- make_emc(data = gng_combined, 
                        design = designGNG_st0, type="single", prior=priorGNG_st0)

# Create the emc objects so we can fit the DDMs (multiple)
sDDM_GNG_SZ_sv <- make_emc(data = gng_combined, 
                           design = designGNG_SZ_sv, type="single", prior=priorGNG_SZ_sv)
sDDM_GNG_SZ_st0 <- make_emc(data = gng_combined, 
                           design = designGNG_SZ_st0, type="single", prior=priorGNG_SZ_st0)
sDDM_GNG_sv_st0 <- make_emc(data = gng_combined, 
                           design = designGNG_sv_st0, type="single", prior=priorGNG_sv_st0)

# Create the emc object for the full DDM
sDDM_GNG_full <- make_emc(data = gng_combined, 
                          design = designGNG_full, type="single", prior=priorGNG_full)
```

### 2.5 Fitting the between-trial variability DDMs

```{r Fitting the between trial variability DDMs}
# Fitting the models
# sDDM_GNG_SZ <- fit(sDDM_GNG_SZ)
# sDDM_GNG_sv <- fit(sDDM_GNG_sv)
# sDDM_GNG_t0 <- fit(sDDM_GNG_t0)
# sDDM_GNG_SZ_sv <- fit(sDDM_GNG_SZ_sv) 
# sDDM_GNG_SZ_st0 <- fit(sDDM_GNG_SZ_st0) 
# sDDM_GNG_sv_st0 <- fit(sDDM_GNG_sv_st0) 
# sDDM_GNG_full <- fit(sDDM_GNG_full)

# Saving the models
# save(sDDM_GNG_SZ, file = here("ddm", "individual", "models", "sDDM_GNG_SZ.Rdata"))
# save(sDDM_GNG_sv, file = here("ddm", "individual", "models", "sDDM_GNG_sv.Rdata"))
# save(sDDM_GNG_t0, file = here("ddm", "individual", "models", "sDDM_GNG_t0.Rdata"))
# save(sDDM_GNG_SZ_sv, file = here("ddm", "individual", "models", "sDDM_GNG_SZ_sv.Rdata"))
# save(sDDM_GNG_SZ_st0, file = here("ddm", "individual", "models", "sDDM_GNG_SZ_st0.Rdata"))
# save(sDDM_GNG_sv_st0, file = here("ddm", "individual", "models", "sDDM_GNG_sv_st0.Rdata"))
# save(sDDM_GNG_full, file = here("ddm", "individual", "models", "sDDM_GNG_full.Rdata"))

# Loading the model objects
load(here("ddm", "individual", "models", "sDDM_GNG_SZ.Rdata"))
load(here("ddm", "individual", "models", "sDDM_GNG_sv.Rdata"))
load(here("ddm", "individual", "models", "sDDM_GNG_t0.Rdata"))
load(here("ddm", "individual", "models", "sDDM_GNG_SZ_sv.Rdata"))
load(here("ddm", "individual", "models", "sDDM_GNG_SZ_st0.Rdata"))
load(here("ddm", "individual", "models", "sDDM_GNG_sv_st0.Rdata"))
load(here("ddm", "individual", "models", "sDDM_GNG_full.Rdata"))
```

### 2.6 Checking model convergence

All models seemd to converge without issues

```{r checking model convergence, warning = F}
check(sDDM_GNG_SZ, plot = FALSE)
check(sDDM_GNG_sv, plot = FALSE)
check(sDDM_GNG_t0, plot = FALSE)
check(sDDM_GNG_SZ_sv, plot = FALSE)
check(sDDM_GNG_SZ_st0, plot = FALSE)
check(sDDM_GNG_sv_st0, plot = FALSE)
check(sDDM_GNG_full, plot = FALSE)
```

## 3. Are our between-trial variability GNG DDMs capturing the data well?

### 3.1 Generating our predicted datasets

```{r Produce predicted values}
# Use the predict function to obtain the fitted values
# ppDDM_GNG_SZ <- predict(sDDM_GNG_SZ, n_cores=4, n_post=200)
# ppDDM_GNG_sv <- predict(sDDM_GNG_sv, n_cores=4, n_post=200)
# ppDDM_GNG_t0 <- predict(sDDM_GNG_t0, n_cores=4, n_post=200)
# ppDDM_GNG_SZ_sv <- predict(sDDM_GNG_SZ_sv, n_cores=4, n_post=200)
# ppDDM_GNG_SZ_st0 <- predict(sDDM_GNG_SZ_st0, n_cores=4, n_post=200)
# ppDDM_GNG_sv_st0 <- predict(sDDM_GNG_sv_st0, n_cores=4, n_post=200)
# ppDDM_GNG_full <- predict(sDDM_GNG_full, n_cores=4, n_post=200)

# Saving the predicted object
# saveRDS(ppDDM_GNG_SZ, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_SZ.rds"))
# saveRDS(ppDDM_GNG_sv, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_sv.rds"))
# saveRDS(ppDDM_GNG_t0, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_t0.rds"))
# saveRDS(ppDDM_GNG_SZ_sv, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_SZ_sv.rds"))
# saveRDS(ppDDM_GNG_SZ_st0, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_SZ_st0.rds"))
# saveRDS(ppDDM_GNG_sv_st0, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_sv_st0.rds"))
# saveRDS(ppDDM_GNG_full, file = here("ddm", "individual", "predicted_values", "ppDDM_GNG_full.rds"))

# Loading the predicted values data frame
ppDDM_GNG_SZ <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_SZ.rds"))
ppDDM_GNG_sv <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_sv.rds"))
ppDDM_GNG_t0 <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_t0.rds"))
ppDDM_GNG_SZ_sv <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_SZ_sv.rds"))
ppDDM_GNG_SZ_st0 <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_SZ_st0.rds"))
ppDDM_GNG_sv_st0 <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_sv_st0.rds"))
ppDDM_GNG_full <- readRDS(here("ddm", "individual", "predicted_values", "ppDDM_GNG_full.rds"))
```

### 3.2 Examining the predicted datasets for the SZ DDM

```{r Examining the predicted datasets for the SZ DDM}
plot_density(gng_combined, ppDDM_GNG_SZ, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_SZ, factors = c("S"))
```

### 3.3 Examining the predicted datasets for the sv DDM

```{r Examining the predicted datasets for the sv DDM}
plot_density(gng_combined, ppDDM_GNG_sv, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_sv, factors = c("S"))
```

### 3.4 Examining the predicted datasets for the t0 DDM

```{r Examining the predicted datasets for the t0 DDM}
plot_density(gng_combined, ppDDM_GNG_t0, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_t0, factors = c("S"))
```

### 3.5 Examining the predicted datasets for the SZ and sv DDM

```{r Examining the predicted datasets for the SZ and sv DDM}
plot_density(gng_combined, ppDDM_GNG_SZ_sv, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_SZ_sv, factors = c("S"))
```

### 3.6 Examining the predicted datasets for the SZ and st0 DDM

```{r Examining the predicted datasets for the SZ and st0 DDM}
plot_density(gng_combined, ppDDM_GNG_SZ_st0, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_SZ_st0, factors = c("S"))
```

### 3.7 Examining the predicted datasets for the sv and st0 DDM

```{r Examining the predicted datasets for the sv and st0 DDM}
plot_density(gng_combined, ppDDM_GNG_sv_st0, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_sv_st0, factors = c("S"))
```

### 3.8 Examining the predicted datasets for the full DDM

```{r Examining the predicted datasets for the full DDM}
plot_density(gng_combined, ppDDM_GNG_full, factors = c("S"))
plot_cdf(gng_combined, ppDDM_GNG_full, factors = c("S"))
```

## 4. Which is the best model?

### 4.1 Model comparison

```{r run a model comparison}
# Load the initial GNG WDM
load(here("ddm", "individual", "models", "sWDM_GNG_combined.Rdata"))

# Run a model comparison
model_comparison <- compare(list(WDM = sWDM_GNG_combined,
                                 SZ=sDDM_GNG_SZ, 
                                 sv=sDDM_GNG_sv, 
                                 t0=sDDM_GNG_t0, 
                                 SZ_sv=sDDM_GNG_SZ_sv,
                                 SZ_st0=sDDM_GNG_SZ_st0,
                                 sv_st0=sDDM_GNG_sv_st0,
                                 full=sDDM_GNG_full))
```
